---
title: "DSC1105_MONFERO_FA3"
author: "John Benedict A. Monfero"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
diamonds
```

### Create a histogram on the diamonds dataset
```{r}
# Let `p` be the container of the histogram visualization below
p <- ggplot(data = diamonds, aes(x = carat))

p # The figure below only shows a blank canvas but the x-axis is all about carat of the diamond dataset
```

### Re-write this using the layer function like we did in class.
```{r}
# Hint: if you don’t know what the default values for some of the aspects of the plot, examine p$layers.
p <- p + layer(
  geom = "bar",
  stat = "bin",
  position = "stack",
  mapping = aes(x = carat),
  params = list(binwidth = 0.1)
)

# Recall p
p # p <- p$layer would give us the proper histogram visualization that we had like in FA1 but with the code: 
  # ggplot(data = diamonds) +
  # geom_histogram(mapping = aes(x = carat))
```
### Remember that a histogram is a plot with `stat_bin` and `geom_bar`.

### Modify your histogram code so that it uses a different `geom`, for example `geom_line` or `geom_point`.
```{r}
# This should be simple once you have the p$layer specification of a histogram.

# let `q` and `r` be the modified histogram codes from p$layer specification with as `geom_line` and `geom_point` respectively.

q <- p + layer(
  geom = "line",
  stat = "bin",
  position = "stack",
  mapping = aes(x = carat),
  params = list(binwidth = 0.1)
)

r <- p + layer(
  geom = "point",
  stat = "bin",
  position = "stack",
  mapping = aes(x = carat),
  params = list(binwidth = 0.1)
)
```

```{r}
# Recall q
q

# Recall r
r
```

### In your histogram in container `p` (the one plotted with bars that you created first):
```{r}
# This is how object `p` visualizes its histogram given the `mapping`, `geom_bar`, `stat_bin`, `position`, and other parameters necessary
p$layers
```

### Add an aesthetic mapping from one of the factor variables (maybe color or clarity) to the fill or color aesthetic.
```{r}
# let `s` and `t` be the modified histogram codes from p$layer but modified according to their aesthetics

s <- ggplot(data = diamonds, aes(x = carat, fill = color)) +
  p$layers

t <- ggplot(data = diamonds, aes(x = carat, fill = clarity)) +
  p$layers
```

```{r}
# Recall s
s
# Recall t
t
```

### What is the default position adjustment for a histogram contained in `s` and `t`? 
```{r}
s$layers
print("As shown above, the default position adjustment for a histogram contained in `s` is `position_stack`")

t$layers
print("As shown above, the default position adjustment for a histogram contained in `t` is `position_stack`")
```

### Try changing the position adjustment in the histogram you created to something different.
```{r}
# If we have complicated layers already, it is also complicated to replicate everything and change only one specific argument to perform the current task!

# We can change some of the properties of s$layer and t$layers without calling all parameters in layer() again:

# Hint: try position = dodge
# Change the position adjustment to dodge
s$layers[[1]]$position <- position_dodge()
t$layers[[1]]$position <- position_dodge()

# Recall s with Updated Position:
s

# Recall s with Updated Position:
t
```

